Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-10 text/manuscript-full-KS.qmd
abstract revisions

*The Structure of College Choice* [@RN4982] created "Geomarkets" and the "Market Segment Model." Geomarkets carve states and metropolitan areas into smaller geographic units, meant to define local recruiting markets. The Market Segment Model predicts how student demand for a particular college varies by Geomarket, based on the socioeconomic characteristics of households. Geomarkets became an input for two College Board products that help colleges recruit students. First, the Enrollment Planning Service (EPS) software recommends specific Geomarkets and high schools from which colleges should recruit. Second, the Student Search Service sells the contact information of prospective students -- referred to as "student lists" -- and colleges can filter by Geomarket to determine which prospect profiles they purchase. We draw from scholarship on quantification, particularly the discussions of correlation and homophily by @RN4975, to conceptualize how recruiting products incorporate Geomarkets. We address two research questions: What is the socioeconomic and racial variation between Geomarkets and how does this variation change over time? How does the socioeconomic and racial composition of included versus excluded prospects vary when student list purchases filter on particular Geomarkets? We answer RQ1 by analyzing Census data from 1980, 2000, and 2020. We answer RQ2 using data on student lists purchased by public universities, which we collected by issuing public records requests. We utilize a quantitative case study design. Metropolitan areas are cases. Analyses consist of descriptive statistics and interactive maps.

@@

The following section introduces background and scholarship about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. Findings for RQ1 reveal that Geomarkets mirror deep-rooted patterns of racial and class segregation, with Black and Latinx communities historically and often still concentrated in the poorest Geomarkets, though these dynamics vary by metropolitan area. Findings for RQ2 indicate student list purchases that exclude working- and middle-class Geomarkets are likely to systematically overlook high-achieving Black and Hispanic prospective students, especially those who are first-generation college students, thereby highlighting how third-party recruitment tools serve as mechanisms of social reproduction by sorting students in ways that reinforce historic inequalities in educational opportunity.

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who developed a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that, "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared [\ldots], big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"

>>>>

*The Structure of College Choice* [@RN4982] created "Geomarkets" and the "Market Segment Mode" (MSM). Geomarkets carve states and metropolitan areas into smaller geographic units to define local recruiting markets. MSM predicts how student demand for a college varies by Geomarket, based on household socioeconomic characteristics. Geomarkets became an input for two College Board recruiting products. First, the Enrollment Planning Service informs which Geomarkets and high schools a college recruits from. Second, the Student Search Service sells prospective student contact information -- or "student lists" -- which colleges filter by Geomarket to determine which prospect profiles to purchase. Drawing from scholarship on quantification, particularly discussions of correlation and homophily, we conceptualize how recruiting products incorporate Geomarkets. We address two research questions: What is the socioeconomic and racial variation between Geomarkets? How does the socioeconomic and racial composition of included versus excluded prospects vary when student lists filter on Geomarkets? We answer RQ1 by analyzing Census data from 1980, 2000, and 2020. RQ2 analyzes data on student lists purchased by public universities, collected through public records requests. Using a quantitative case study design with metropolitan areas as cases, descriptive statistics and interactive maps suggest Geomarkets mirror deeply-rooted patterns of racial and class segregation. Moreover, student lists that exclude low-income Geomarkets systematically overlook non-white students, first-generation students, and especially first-generation non-white students.

@@

The following section introduces background and scholarship about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. Findings for RQ1 reveal that Geomarkets mirror deep-rooted patterns of racial and class segregation, with Black and Latinx communities historically and often still concentrated in the poorest Geomarkets, though these dynamics vary by metropolitan area. Findings for RQ2 indicate student list purchases that exclude low-income Geomarkets are likely to systematically overlook high-achieving Black, Hispanic, and Asian prospective students, especially those who are first-generation college students, thereby highlighting how third-party recruitment tools serve as mechanisms of social reproduction by sorting students in ways that reinforce historic inequalities in educational opportunity.

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who developed a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that, "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \[\ldots\], big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"
